---
title: "TabulaMuris_Liver_Bootstrapping"
output:
  html_document:
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(
  dpi = 300,
  fig.width = 5,   # 1500 pixels / 300 dpi
  fig.height = 4   # 1200 pixels / 300 dpi
)
```

#Install R packages
```{r install-packages}
pkgs <- c("here", "ggplot2", "dplyr", "Seurat", "skmeans", "mclust", "tidyr", "ggrepel", "stringr", "igraph")
to_install <- setdiff(pkgs, rownames(installed.packages()))
if (length(to_install)) {
  install.packages(to_install)
}
```

#OPTIMAL PC SELECTION ALGO
```{r quick-elbow}
quick.elbow <- function(varpc,low=.08,max.pc=.9) {
  ee <- varpc/sum(varpc) #proportinal eigenvalues
  while(low>=max(ee)) { low <- low/2 } #when no principal component eigenvalues are large enough, cut low in half
  lowie <- (ee<low) ; highie <- ee>low/8
  low.ones <- which(lowie & highie)
  others <- length(which(!lowie))
  if(length(low.ones)>0) {
    if(length(low.ones)==1) {
      elbow <- low.ones
    } else {
      set <- ee[low.ones]
      pc.drops <- abs(diff(set))/(set[1:(length(set)-1)])
      infz <- is.infinite(pc.drops)
      #print(pc.drops)
      elbow <- which(pc.drops==max(pc.drops[!infz],na.rm=T))[1]+others
    }
  } else {
    # if somehow there are no small eigenvalues, just choose the elbow as the second last
    cat("no eigenvalues were significantly smaller than the previous\n")
    elbow <- length(ee)
  }
  if(tail(cumsum(ee[1:elbow]),1)>max.pc) {
    elbow <- which(cumsum(ee)>max.pc)[1]-1
  }
  if(elbow<1) {
    warning("elbow calculation failed, return zero")
    return(0)
  }
  elbow<-elbow + 1 # include last to be sure ..
  names(elbow) <- NULL
  return(elbow)
}
```

#GENES FILTERING
```{r gene-filter}
library(stringr)
genes_filter <- function(df, filters = c("IgFilter", "TrFilter", "riboFilter", "mtFilter", "gmFilter", "rikFilter", "dotFilter")) {
  #ensure gene names are uppercase for consistent matching
  geneNames <- toupper(rownames(df))
  GenesToFilterOut <- integer(0)
  if(is.null(filters)){
    return(df)
  }

  #immunoglobulin genes (IGHV, IGKV)
  if ("IgFilter" %in% filters) {
    GenesToFilterOut <- c(
      GenesToFilterOut,
      which(str_starts(geneNames, "IGHV") | str_starts(geneNames, "IGKV"))
    )
  }

  #t-cell receptor genes (TRAV, TRBV)
  if ("TrFilter" %in% filters) {
    GenesToFilterOut <- c(
      GenesToFilterOut,
      which(str_starts(geneNames, "TRAV") | str_starts(geneNames, "TRBV"))
    )
  }

  #ribosomal genes (RPL, RPS)
  if ("riboFilter" %in% filters) {
    GenesToFilterOut <- c(
      GenesToFilterOut,
      which(str_starts(geneNames, "RPL") | str_starts(geneNames, "RPS"))
    )
  }

  #mitochondrial genes (MT-)
  if ("mtFilter" %in% filters) {
    GenesToFilterOut <- c(
      GenesToFilterOut,
      which(str_starts(geneNames, "MT-"))
    )
  }

  #GM-prefixed genes
  if ("gmFilter" %in% filters) {
    GenesToFilterOut <- c(
      GenesToFilterOut,
      which(str_starts(geneNames, "GM"))
    )
  }

  #genes ending with RIK
  if ("rikFilter" %in% filters) {
    GenesToFilterOut <- c(
      GenesToFilterOut,
      which(str_ends(geneNames, "RIK"))
    )
  }

  #genes starting with "\\.", ENSG, or a regex pattern like XX000000
  if ("dotFilter" %in% filters) {
    GenesToFilterOut <- c(
      GenesToFilterOut,
      which(
        str_starts(geneNames, "\\.") |
        str_starts(geneNames, "ENSG") |
        str_detect(geneNames, "^[A-Z]{2}\\d{6}")
      )
    )
  }

  #remove duplicate indices
  GenesToFilterOut <- unique(GenesToFilterOut)

  #return filtered dataframe
  if (length(GenesToFilterOut) > 0) {
    filtered_df <- df[-GenesToFilterOut, , drop = FALSE]
  } else {
    filtered_df <- df  # No genes to filter, return original
  }

  list(
    counts=filtered_df,
    filtered=setdiff(rownames(df), rownames(filtered_df))
  )
}
```

#PRINCIPAL GENES ALGO
```{r main-algo}
principalgenes<- function(data = NULL, pcs = -1, cut = -1, filters=c("IgFilter", "TrFilter", "riboFilter", "mtFilter", "gmFilter", "rikFilter", "dotFilter")) {

  message(length(rownames(data)), ", ", length(colnames(data)))
  message(paste0(filters))
  if (is.null(data)) {
    stop("Error: 'data' must be provided.")
  }

  if(cut == -1){
    cut = 0.25
  }
  
  message("filtering genes based on specifications")
  data<-genes_filter(count_matrix, filters)
  filtered_genes<-data$filtered
  data<-data$counts
  message("Post Filtering Dataset Dimensions: Rows: ", length(rownames(data)), ", Columns: ", length(colnames(data)))
  
  zero_cols<-which(colSums(data) == 0)
  zero_var_rows <- apply(data, 1, var, na.rm = TRUE) == 0
  zero_var_rows<-which(zero_var_rows)

  message("Remove zero rows and remove zero columns until none are left")
  while(length(zero_cols) != 0 || length(zero_var_rows)!=0){
    zero_cols<-which(colSums(data) == 0)
    if(length(zero_cols) > 0) {
      data<-data[, -zero_cols]
    }
    zero_var_rows <- apply(data, 1, var, na.rm = TRUE) == 0
    zero_var_rows<-which(zero_var_rows)
    if(length(zero_var_rows) > 0) {
      data<-data[-zero_var_rows, ]
    }
    message("Dataset Dimensions after removing zero rows and columns: Rows: ", length(rownames(data)), ", Columns: ", length(colnames(data)))
  }
  
  message("Calculating optimal number of PCs via elbow method:")
  if(pcs == -1 || pcs >= 20 || pcs >= 0.05*ncol(data)){
    temp<- irlba::prcomp_irlba(t(data), n = min(20, 0.05*ncol(data)), retx = TRUE, center = TRUE, scale. = TRUE)
    eig = temp$sdev^2
    pcs = quick.elbow(eig)
    rm(eig, temp)
  }

  message("Values set: pcs = ", pcs, ", chosen cutoff = ", cut)

  #transpose the data to have genes as columns and cells as rows
  rotated_output <- irlba::prcomp_irlba(t(data), n = pcs, retx = TRUE, center = TRUE, scale. = TRUE)

  #prepare to represent results
  summary<- summary(rotated_output)
  loadings <- rotated_output$rotation
  rownames(loadings) <- rownames(data) # Associate gene names with loadings

  #compute variance explained by each PC
  eigenvalues <- (rotated_output$sdev)^2
  prop_var <- eigenvalues / sum(eigenvalues)
  temp_obj <- data.frame(x=colnames(loadings), y = prop_var)

  #plot PCA results
  graph1<- ggplot2::ggplot(temp_obj, ggplot2::aes(x = x, y = y * 100)) +
    ggplot2::geom_bar(stat = "identity", fill = "skyblue") +
    ggplot2::geom_text(ggplot2::aes(label= round(y*100, 2)), vjust = +1.5)+
    ggplot2::labs(x = NULL, y = NULL, title = NULL) +
    ggplot2::theme_minimal() + ggplot2::theme(legend.position = "none")
  plot(graph1)

  #calculate absolute values of loadings and contribution percentages
  loadings <- abs(loadings)
  sumload <- colSums(loadings)
  percent_cont <- loadings / sumload

  #calculate total contribution to variance for each gene
  scaled_contr<-sweep(percent_cont, 2, prop_var, FUN = "*")
  tot_contr <- rowSums(scaled_contr)
  
  #select genes with total contribution in cut percentile
  message("starting the cutoff process")
  cumulative_gps<- 0
  if (cut == 1){
    result<- sort(decreasing = TRUE, tot_contr)
  } else {

    for (i in 1:length(tot_contr)){
      result<-list()
      if (is.na(tot_contr[i]) || is.infinite(tot_contr[i])){
        stop("Error: contribution calculation produced NA or infinite values. Please check the input data.")
      }

      result[[i]] <- as.numeric(tot_contr[i])
      cumulative_gps <- cumulative_gps + as.numeric(tot_contr[i])
      if (cumulative_gps >= cut){
        break
      }
    }
    result<-unlist(result)
  }

  # Log transformation for top genes' contributions
  logcont <- -10 * log10(result)

  # Return results
  list(
    summary,
    log_contributions = logcont, #-10*log10(gps), larger values imply less contribution
    top_genes = result, #ordered top selected genes
    rotated_output = rotated_output, #pca output
    filtered = filtered_genes, #list of genes filtered out
    count_matrix = data #filtered count matrix for use elsewhere
  )

}
```

#Global Library
```{r here-library}
library(here)
```

#TabulaMuris_Liver DATASET
##Read in the dataset
```{r read-data}

input_file<-paste0(here(), "/Filtered_Input/TabulaMuris_Liver_Counts.csv")
truth_file<-paste0(here(), "/Filtered_Input/TabulaMuris_Liver_Labels.csv")
name<-"TabulaMuris_Liver"

count_matrix<-read.csv(input_file, row.names = 1)
rownames(count_matrix)<-gsub("-", "_", rownames(count_matrix)) # consistent naming purposes
colnames(count_matrix)<-gsub("\\.", "-", colnames(count_matrix))
```

##Get principal genes and highlight them on the log_var_mean_plot
```{r get-principal-genes}

#plot variance vs mean expression
library(ggplot2)
library(dplyr)

row_means<-rowMeans(count_matrix, na.rm=TRUE)
row_var<- apply(count_matrix, 1, var, na.rm=TRUE)

log_var_mean_df<-tibble(
  gene_id=rownames(count_matrix),
  log_mean = log10(row_means +  1),
  log_var = log10(row_var + 1)
)

#get top principal genes
set.seed(12345)
principal_genes_results <- principalgenes(count_matrix, cut = 1, filters = c("riboFilter", "mtFilter", "gmFilter", "rikFilter"))
filtered_counts<-principal_genes_results$count_matrix
top_50_genes<-names(principal_genes_results$top_genes[1:50])

#highlight top genes selected by principal genes method
log_var_mean_df<- log_var_mean_df %>% mutate(highlight = gene_id %in% top_50_genes)
```

##PCA PLOT
```{r pca-plot}
PC1_and_PC2<-as.data.frame(principal_genes_results$rotated_output$x, row.names = colnames(filtered_counts))

#get true labels
full_true_labels<-read.csv(truth_file)
full_true_labels$cell<-gsub("\\.", "-", full_true_labels$cell)
full_true_labels$numeric_labels<-as.numeric(factor(
  full_true_labels$label,
  levels = unique(full_true_labels$label)
))
true_labels<-subset(full_true_labels, cell %in% colnames(filtered_counts))
PC1_and_PC2$labels<-true_labels$label
PC1_and_PC2$numeric_labels <- as.numeric(factor(
  PC1_and_PC2$labels,
  levels = unique(true_labels$label)
))

#PCA PLOT
ggplot(PC1_and_PC2, aes(x=PC1, y=PC2, color=labels))+
  geom_point(size=3)+
  scale_color_brewer(palette="Set1",
                     name = "Label") +
  theme_minimal() + theme(legend.position = "none")+
  labs(x = NULL, y = NULL, title = NULL)+
  theme(plot.title=element_text(hjust=0.5))

ggsave(filename = paste0(here(), "/Bootstrapping_Results/TabulaMuris_Liver_Results/TabulaMuris_Liver_PCA_Plot.jpg"),, 
    dpi = 300)
```

##BUILD CUMULATIVE GENE LISTS BY THRESHOLD
```{r build-cumulative-gene-lists}
source(paste0(here(), "/clSensitivity.R"))
library(skmeans)
library(mclust)
library(igraph)
library(dplyr)
library(ggplot2)

thresholds <- seq(10, 50, by = 2)

selectedGeneList <- list()
current_gene_list <- list()
currentpropsum <- 0
threshold_index <- 1

for(i in seq_along(principal_genes_results$top_genes)){
  
  current_gene_list <- append(current_gene_list, principal_genes_results$top_genes[i])
  
  currentpropsum <- currentpropsum + principal_genes_results$top_genes[i] * 100
  
  if(currentpropsum >= thresholds[threshold_index]){
    selectedGeneList[[threshold_index]] <- current_gene_list
    
    print(paste("Threshold", thresholds[threshold_index], "reached at index", i))
    
    threshold_index <- threshold_index + 1
    
    if(threshold_index > length(thresholds)){
      break
    }
  }
}

num_genes <- list()
for (i in seq_along(selectedGeneList)){
  num_genes[[i]]<-length(selectedGeneList[[i]])
}
num_genes<-unlist(num_genes)

true_labels_vector <- setNames(PC1_and_PC2$numeric_labels, row.names(PC1_and_PC2))
k_cls <- max(true_labels_vector)
```

##BUILD HVG GENE LISTS (SEURAT), MATCHED TO THE SAME THRESHOLDS AS PG
```{r build-hvg-gene-lists}
library(Seurat)

seuratObject <- CreateSeuratObject(counts = as.matrix(count_matrix))
Idents(seuratObject) <- "All_Cells"
norm_seuratObject <- NormalizeData(seuratObject)
seuratObject <- FindVariableFeatures(norm_seuratObject, selection.method = "dispersion", nfeatures = length(rownames(count_matrix))) #dispersion value = log(variance/mean)

scored_hvg <- seuratObject[["RNA"]]@meta.data$vf_disp_data_mvp.dispersion
names(scored_hvg) <- rownames(count_matrix) #restore original gene names (order-preserved)
scored_hvg[is.na(scored_hvg)] <- 0
scored_hvg <- sort(decreasing = TRUE, scored_hvg)
HVGs <- names(scored_hvg) #descending-order HVG gene names, aligned with count_matrix rownames

# Build HVG gene lists sized to match the number of genes selected by PG at each threshold
HVGGeneList <- list()
for (j in seq_along(num_genes)) {
  HVGGeneList[[j]] <- HVGs[1:num_genes[j]]
}
```

##BUILD UNION (PG + HVG) GENE LISTS AT EACH THRESHOLD
```{r build-union-gene-lists}
UnionGeneList <- list()
for (j in seq_along(selectedGeneList)) {
  UnionGeneList[[j]] <- unique(union(names(selectedGeneList[[j]]), HVGs[1:num_genes[j]]))
}
union_num_genes <- as.integer(sapply(UnionGeneList, length))
```

##SK_MEANS CLUSTERING ON RANDOM 80% CELL SUBSETS FOR EACH GENE-SELECTION METHOD
```{r sk-means-subsampled-accuracy}
n_subsets <- 5
subset_frac <- 0.8
seeds <- c(12345, 54672, 44182, 72198, 98230)

all_cells <- colnames(count_matrix)
subset_size <- floor(subset_frac * length(all_cells))

# Matrices to hold accuracy: rows = subsets, columns = threshold levels
accuracy_matrix_pg    <- matrix(NA_real_, nrow = n_subsets, ncol = length(selectedGeneList))
accuracy_matrix_hvg   <- matrix(NA_real_, nrow = n_subsets, ncol = length(HVGGeneList))
accuracy_matrix_union <- matrix(NA_real_, nrow = n_subsets, ncol = length(UnionGeneList))

for (r in 1:n_subsets){
  set.seed(seeds[r])
  cell_subset <- sample(all_cells, subset_size)
  sub_true_labels <- true_labels_vector[cell_subset]

  print(paste("subset:", r))
  for (j in seq_along(selectedGeneList)){
    # Principal Genes
    submat_pg <- count_matrix[names(selectedGeneList[[j]]), cell_subset]
    set.seed(seeds[r])
    sub_cluster_pg <- skmeans(x = t(log2(submat_pg + 1)), k = k_cls, control = list(verbose = FALSE))$cluster
    sub_sens_pg <- clustSensitivity(sub_cluster_pg, sub_true_labels, as.character(1:k_cls))
    accuracy_matrix_pg[r, j] <- as.numeric(sub_sens_pg[1, 2])

    # Seurat HVGs
    submat_hvg <- count_matrix[HVGGeneList[[j]], cell_subset]
    set.seed(seeds[r])
    sub_cluster_hvg <- skmeans(x = t(log2(submat_hvg + 1)), k = k_cls, control = list(verbose = FALSE))$cluster
    sub_sens_hvg <- clustSensitivity(sub_cluster_hvg, sub_true_labels, as.character(1:k_cls))
    accuracy_matrix_hvg[r, j] <- as.numeric(sub_sens_hvg[1, 2])

    # PG / HVG Union
    submat_union <- count_matrix[UnionGeneList[[j]], cell_subset]
    set.seed(seeds[r])
    sub_cluster_union <- skmeans(x = t(log2(submat_union + 1)), k = k_cls, control = list(verbose = FALSE))$cluster
    sub_sens_union <- clustSensitivity(sub_cluster_union, sub_true_labels, as.character(1:k_cls))
    accuracy_matrix_union[r, j] <- as.numeric(sub_sens_union[1, 2])

    print(paste0(j, "th threshold clustering done"))
  }
}

summarize_accuracy <- function(accuracy_matrix, n_subsets) {
  mean_accuracy   <- colMeans(accuracy_matrix)
  sd_accuracy     <- apply(accuracy_matrix, 2, sd)
  se_accuracy     <- sd_accuracy / sqrt(n_subsets)
  margin_of_error <- qt(0.975, df = n_subsets - 1) * se_accuracy
  data.frame(
    mean_accuracy   = as.numeric(mean_accuracy),
    sd_accuracy     = as.numeric(sd_accuracy),
    margin_of_error = as.numeric(margin_of_error),
    ci_lower        = as.numeric(mean_accuracy - margin_of_error),
    ci_upper        = as.numeric(mean_accuracy + margin_of_error)
  )
}

pg_summary    <- summarize_accuracy(accuracy_matrix_pg, n_subsets)
hvg_summary   <- summarize_accuracy(accuracy_matrix_hvg, n_subsets)
union_summary <- summarize_accuracy(accuracy_matrix_union, n_subsets)

points_sk <- data.frame(
  thresholds        = as.numeric(thresholds),
  num_genes         = as.integer(num_genes),
  union_num_genes   = as.integer(union_num_genes),
  PG_mean           = pg_summary$mean_accuracy,
  PG_ci_lower       = pg_summary$ci_lower,
  PG_ci_upper       = pg_summary$ci_upper,
  HVG_mean          = hvg_summary$mean_accuracy,
  HVG_ci_lower      = hvg_summary$ci_lower,
  HVG_ci_upper      = hvg_summary$ci_upper,
  Union_mean        = union_summary$mean_accuracy,
  Union_ci_lower    = union_summary$ci_lower,
  Union_ci_upper    = union_summary$ci_upper
)

# Long (tidy) format for ggplot
plot_df1 <- bind_rows(
  data.frame(thresholds = points_sk$thresholds, num_genes = points_sk$num_genes,
             union_num_genes = points_sk$union_num_genes,
             mean_accuracy = points_sk$PG_mean, ci_lower = points_sk$PG_ci_lower,
             ci_upper = points_sk$PG_ci_upper, metric = "PG_sensitivity"),
  data.frame(thresholds = points_sk$thresholds, num_genes = points_sk$num_genes,
             union_num_genes = points_sk$union_num_genes,
             mean_accuracy = points_sk$HVG_mean, ci_lower = points_sk$HVG_ci_lower,
             ci_upper = points_sk$HVG_ci_upper, metric = "HVG_sensitivity"),
  data.frame(thresholds = points_sk$thresholds, num_genes = points_sk$num_genes,
             union_num_genes = points_sk$union_num_genes,
             mean_accuracy = points_sk$Union_mean, ci_lower = points_sk$Union_ci_lower,
             ci_upper = points_sk$Union_ci_upper, metric = "PG_HVG_Union")
)

#control legend order
plot_df1$metric <- factor(plot_df1$metric,
                           levels = c("PG_sensitivity", "HVG_sensitivity", "PG_HVG_Union"))

#add labels for gene-count annotations
plot_df1$label_val <- NA
plot_df1$label_val[plot_df1$metric == "PG_sensitivity"] <- plot_df1$num_genes[plot_df1$metric == "PG_sensitivity"]
plot_df1$label_val[plot_df1$metric == "PG_HVG_Union"] <- plot_df1$union_num_genes[plot_df1$metric == "PG_HVG_Union"]

write.csv(plot_df1, file = paste0(here(), "/Bootstrapping_Data_Hold/TabulaMuris_Liver_PG_HVG_Union_CumulativeScore_Subsampled_Accuracy.csv"))
```

##PLOT MEAN CLUSTERING SENSITIVITY VS CUMULATIVE SCORE WITH 95% CI (PG, HVG, UNION)
```{r plot-subsampled-accuracy}
ggplot(plot_df1, aes(x = thresholds, y = mean_accuracy, color = metric)) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.5) +

  # PG labels
  geom_text(
    data = subset(plot_df1, metric == "PG_sensitivity"),
    aes(label = label_val),
    color = "#1BB7D8", vjust = +2, size = 3, check_overlap = TRUE
  ) +

  # Union labels
  geom_text(
    data = subset(plot_df1, metric == "PG_HVG_Union"),
    aes(label = label_val),
    color = "darkgreen", vjust = -2, size = 3, check_overlap = TRUE
  ) +

  geom_smooth(
    method = "loess",
    method.args = list(span = 1, degree = 1, surface = "direct"),
    se = FALSE
  ) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = NULL, y = NULL, title = NULL) +
  scale_color_manual(
    breaks = c("PG_sensitivity", "HVG_sensitivity", "PG_HVG_Union"),
    values = c(
      "PG_sensitivity"  = "#1BB7D8",
      "HVG_sensitivity" = "#E55B1E",
      "PG_HVG_Union"    = "#2C9F45"
    ),
    labels = c(
      "PG_sensitivity"  = "Principal Genes",
      "HVG_sensitivity" = "Seurat's HVG",
      "PG_HVG_Union"    = "HVG_PG_Union"
    )
  ) +
  theme_minimal() + theme(legend.position = "none")

ggsave(filename = paste0(here(), "/Bootstrapping_Results/TabulaMuris_Liver_Results/TabulaMuris_Liver_PG_HVG_Union_CumulativeScore_Subsampled_Sensitivity.jpg"),, 
     dpi = 300)
```