---
title: "Vst_TabulaMuris_Liver"
output:
  html_document:
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(
  dpi = 300,
  fig.width = 5,   # 1500 pixels / 300 dpi
  fig.height = 4   # 1200 pixels / 300 dpi
)
```

#Install R packages
```{r}
pkgs <- c("here", "ggplot2", "dplyr", "Seurat", "skmeans", "mclust", "tidyr", "ggrepel", "stringr", "igraph")
to_install <- setdiff(pkgs, rownames(installed.packages()))
if (length(to_install)) {
  install.packages(to_install)
}
```

#OPTIMAL PC SELECTION ALGO
```{r quick-elbow}
quick.elbow <- function(varpc,low=.08,max.pc=.9) {
  ee <- varpc/sum(varpc) #proportinal eigenvalues
  while(low>=max(ee)) { low <- low/2 } #when no principal component eigenvalues are large enough, cut low in half
  lowie <- (ee<low) ; highie <- ee>low/8
  low.ones <- which(lowie & highie)
  others <- length(which(!lowie))
  if(length(low.ones)>0) {
    if(length(low.ones)==1) {
      elbow <- low.ones
    } else {
      set <- ee[low.ones]
      pc.drops <- abs(diff(set))/(set[1:(length(set)-1)])
      infz <- is.infinite(pc.drops)
      #print(pc.drops)
      elbow <- which(pc.drops==max(pc.drops[!infz],na.rm=T))[1]+others
    }
  } else {
    # if somehow there are no small eigenvalues, just choose the elbow as the second last
    cat("no eigenvalues were significantly smaller than the previous\n")
    elbow <- length(ee)
  }
  if(tail(cumsum(ee[1:elbow]),1)>max.pc) {
    elbow <- which(cumsum(ee)>max.pc)[1]-1
  }
  if(elbow<1) {
    warning("elbow calculation failed, return zero")
    return(0)
  }
  elbow<-elbow + 1 # include last to be sure ..
  names(elbow) <- NULL
  return(elbow)
}
```

#GENES FILTERING
```{r gene-filter}
library(stringr)
genes_filter <- function(df, filters = c("IgFilter", "TrFilter", "riboFilter", "mtFilter", "gmFilter", "rikFilter", "dotFilter")) {
  #ensure gene names are uppercase for consistent matching
  geneNames <- toupper(rownames(df))
  GenesToFilterOut <- integer(0)
  if(is.null(filters)){
    return(df)
  }

  #immunoglobulin genes (IGHV, IGKV)
  if ("IgFilter" %in% filters) {
    GenesToFilterOut <- c(
      GenesToFilterOut,
      which(str_starts(geneNames, "IGHV") | str_starts(geneNames, "IGKV"))
    )
  }

  #t-cell receptor genes (TRAV, TRBV)
  if ("TrFilter" %in% filters) {
    GenesToFilterOut <- c(
      GenesToFilterOut,
      which(str_starts(geneNames, "TRAV") | str_starts(geneNames, "TRBV"))
    )
  }

  #ribosomal genes (RPL, RPS)
  if ("riboFilter" %in% filters) {
    GenesToFilterOut <- c(
      GenesToFilterOut,
      which(str_starts(geneNames, "RPL") | str_starts(geneNames, "RPS"))
    )
  }

  #mitochondrial genes (MT-)
  if ("mtFilter" %in% filters) {
    GenesToFilterOut <- c(
      GenesToFilterOut,
      which(str_starts(geneNames, "MT-"))
    )
  }

  #GM-prefixed genes
  if ("gmFilter" %in% filters) {
    GenesToFilterOut <- c(
      GenesToFilterOut,
      which(str_starts(geneNames, "GM"))
    )
  }

  #genes ending with RIK
  if ("rikFilter" %in% filters) {
    GenesToFilterOut <- c(
      GenesToFilterOut,
      which(str_ends(geneNames, "RIK"))
    )
  }

  #genes starting with "\\.", ENSG, or a regex pattern like XX000000
  if ("dotFilter" %in% filters) {
    GenesToFilterOut <- c(
      GenesToFilterOut,
      which(
        str_starts(geneNames, "\\.") |
        str_starts(geneNames, "ENSG") |
        str_detect(geneNames, "^[A-Z]{2}\\d{6}")
      )
    )
  }

  #remove duplicate indices
  GenesToFilterOut <- unique(GenesToFilterOut)

  #return filtered dataframe
  if (length(GenesToFilterOut) > 0) {
    filtered_df <- df[-GenesToFilterOut, , drop = FALSE]
  } else {
    filtered_df <- df  # No genes to filter, return original
  }

  list(
    counts=filtered_df,
    filtered=setdiff(rownames(df), rownames(filtered_df))
  )
}
```

#PRINCIPAL GENES ALGO
```{r main-algo}
principalgenes<- function(data = NULL, pcs = -1, cut = -1, filters=c("IgFilter", "TrFilter", "riboFilter", "mtFilter", "gmFilter", "rikFilter", "dotFilter")) {

  message(length(rownames(data)), ", ", length(colnames(data)))
  message(paste0(filters))
  if (is.null(data)) {
    stop("Error: 'data' must be provided.")
  }

  if(cut == -1){
    cut = 0.25
  }
  
  message("filtering genes based on specifications")
  data<-genes_filter(count_matrix, filters)
  filtered_genes<-data$filtered
  data<-data$counts
  message("Post Filtering Dataset Dimensions: Rows: ", length(rownames(data)), ", Columns: ", length(colnames(data)))
  
  zero_cols<-which(colSums(data) == 0)
  zero_var_rows <- apply(data, 1, var, na.rm = TRUE) == 0
  zero_var_rows<-which(zero_var_rows)

  message("Remove zero rows and remove zero columns until none are left")
  while(length(zero_cols) != 0 || length(zero_var_rows)!=0){
    zero_cols<-which(colSums(data) == 0)
    if(length(zero_cols) > 0) {
      data<-data[, -zero_cols]
    }
    zero_var_rows <- apply(data, 1, var, na.rm = TRUE) == 0
    zero_var_rows<-which(zero_var_rows)
    if(length(zero_var_rows) > 0) {
      data<-data[-zero_var_rows, ]
    }
    message("Dataset Dimensions after removing zero rows and columns: Rows: ", length(rownames(data)), ", Columns: ", length(colnames(data)))
  }
  
  message("Calculating optimal number of PCs via elbow method:")
  if(pcs == -1 || pcs >= 20 || pcs >= 0.05*ncol(data)){
    temp<- irlba::prcomp_irlba(t(data), n = min(20, 0.05*ncol(data)), retx = TRUE, center = TRUE, scale. = TRUE)
    eig = temp$sdev^2
    pcs = quick.elbow(eig)
    rm(eig, temp)
  }

  message("Values set: pcs = ", pcs, ", chosen cutoff = ", cut)

  #transpose the data to have genes as columns and cells as rows
  rotated_output <- irlba::prcomp_irlba(t(data), n = pcs, retx = TRUE, center = TRUE, scale. = TRUE)

  #prepare to represent results
  summary<- summary(rotated_output)
  loadings <- rotated_output$rotation
  rownames(loadings) <- rownames(data) # Associate gene names with loadings

  #compute variance explained by each PC
  eigenvalues <- (rotated_output$sdev)^2
  prop_var <- eigenvalues / sum(eigenvalues)
  temp_obj <- data.frame(x=colnames(loadings), y = prop_var)

  #plot PCA results
  graph1<- ggplot2::ggplot(temp_obj, ggplot2::aes(x = x, y = y * 100)) +
    ggplot2::geom_bar(stat = "identity", fill = "skyblue") +
    ggplot2::geom_text(ggplot2::aes(label= round(y*100, 2)), vjust = +1.5)+
    ggplot2::labs(x = NULL, y = NULL, title = NULL) +
    ggplot2::theme_minimal() + ggplot2::theme(legend.position = "none")
  plot(graph1)

  #calculate absolute values of loadings and contribution percentages
  loadings <- abs(loadings)
  sumload <- colSums(loadings)
  percent_cont <- loadings / sumload

  #calculate total contribution to variance for each gene
  scaled_contr<-sweep(percent_cont, 2, prop_var, FUN = "*")
  tot_contr <- rowSums(scaled_contr)
  
  #select genes with total contribution in cut percentile
  message("starting the cutoff process")
  cumulative_gps<- 0
  if (cut == 1){
    result<- sort(decreasing = TRUE, tot_contr)
  } else {

    for (i in 1:length(tot_contr)){
      result<-list()
      if (is.na(tot_contr[i]) || is.infinite(tot_contr[i])){
        stop("Error: contribution calculation produced NA or infinite values. Please check the input data.")
      }

      result[[i]] <- as.numeric(tot_contr[i])
      cumulative_gps <- cumulative_gps + as.numeric(tot_contr[i])
      if (cumulative_gps >= cut){
        break
      }
    }
    result<-unlist(result)
  }

  # Log transformation for top genes' contributions
  logcont <- -10 * log10(result)

  # Return results
  list(
    summary,
    log_contributions = logcont, #-10*log10(gps), larger values imply less contribution
    top_genes = result, #ordered top selected genes
    rotated_output = rotated_output, #pca output
    filtered = filtered_genes, #list of genes filtered out
    count_matrix = data #filtered count matrix for use elsewhere
  )

}
```

#Global Library
```{r}
library(here)
```

#TABULA_MURIS_LIVER

```{r}

input_file<-"/Filtered_Input/TabulaMuris_Liver_Counts.csv"
truth_file<-"/Filtered_Input/TabulaMuris_Liver_Labels.csv"
name<-"TabulaMuris_Liver"

count_matrix<-read.csv(paste0(here(), input_file), check.names = FALSE, row.names = 1)
full_true_labels<-read.csv(paste0(here(), truth_file), check.names = FALSE)
```

##Get principal genes and highlight them on the log_var_mean_plot
```{r}

#compare variance vs mean expression
library(ggplot2)
library(dplyr)

row_means<-rowMeans(count_matrix, na.rm=TRUE)
row_var<- apply(count_matrix, 1, var, na.rm=TRUE)

log_var_mean_df<-tibble(
  gene_id=rownames(count_matrix),
  log_mean = log10(row_means +  1),
  log_var = log10(row_var + 1)
)

set.seed(12345)
principal_genes_results <- principalgenes(count_matrix, cut = 1, filters=c("riboFilter", "mtFilter", "gmFilter", "rikFilter"))
filtered_counts<-principal_genes_results$count_matrix
top_100_genes<-names(principal_genes_results$top_genes[1:50])

```
##PCA PLOT
```{r}
PC1_and_PC2<-as.data.frame(principal_genes_results$rotated_output$x, row.names = colnames(filtered_counts))

#get true labels
full_true_labels$numeric_labels<-as.numeric(factor(
  full_true_labels$label,
  levels = unique(full_true_labels$label)
))
true_labels<-subset(full_true_labels, cell %in% colnames(filtered_counts))
PC1_and_PC2$labels<-true_labels$label
PC1_and_PC2$numeric_labels <- as.numeric(factor(
  PC1_and_PC2$labels,
  levels = unique(true_labels$label)
))

#PCA PLOT
ggplot(PC1_and_PC2, aes(x=PC1, y=PC2, color=labels))+
  geom_point(size=3)+
  scale_color_brewer(palette="Set1",
                     name = "Label") +
  theme_minimal() + theme(legend.position = "none")+
  labs(x = NULL, y = NULL, title = NULL)+
  theme(plot.title=element_text(hjust=0.5))
ggsave(filename = paste0(here(), "/vst_results/TabulaMuris_Liver_Results/TabulaMuris_Liver_PCA_Plot.jpg"),
        dpi = 300)
```

##SK_MEANS CLUSTERING USING CUMULATIVE SCORE
```{r}
source(paste0(here(), "/clSensitivity.R"))
library(skmeans)
library(igraph)

thresholds <- seq(10, 50, by = 2)

selectedGeneList <- list()
current_gene_list <- list()
currentpropsum <- 0
threshold_index <- 1

for(i in seq_along(principal_genes_results$top_genes)){
  
  current_gene_list <- append(current_gene_list, principal_genes_results$top_genes[i])
  
  currentpropsum <- currentpropsum + principal_genes_results$top_genes[i] * 100
  
  if(currentpropsum >= thresholds[threshold_index]){
    selectedGeneList[[threshold_index]] <- current_gene_list
    
    print(paste("Threshold", thresholds[threshold_index], "reached at index", i))
    
    threshold_index <- threshold_index + 1
    
    if(threshold_index > length(thresholds)){
      break
    }
  }
}

num_genes <- list()
for (i in seq_along(selectedGeneList)){
  num_genes[[i]]<-length(selectedGeneList[[i]])
}
num_genes<-unlist(num_genes)

sumprop<- 0
for (i in seq_along(principal_genes_results$top_genes)){
  sumprop <- sumprop + principal_genes_results$top_genes[i]*100
}

true_labels_vector <- setNames(PC1_and_PC2$numeric_labels, row.names(PC1_and_PC2))

k_cls <- max(true_labels_vector)
# uncomment the line below to select specific number of clusters
# k_cls <- 3
PG_groups_sk <- list()
PG_clustering_sk <- list()
PG_results_sk <-list()
PG_clustering_sens_sk<-list()
seeds<-c(12345, 54672, 44182, 72198, 98230)

for (i in 1:length(seeds)){
  print(paste("seed:", seeds[i]))
  for (j in 1:length(selectedGeneList)){
    set.seed(seeds[i])
    submat<-count_matrix[names(selectedGeneList[[j]]),]
    PG_groups_sk[[j]]<-skmeans(x = t(log2(submat + 1)) , k = k_cls, control = list(verbose = FALSE))$cluster
    print(paste0(j, "th clustering done"))
  }
  PG_clustering_sk[[i]]<-PG_groups_sk
}

for (i in 1:length(PG_clustering_sk)){
  print(paste("seed:", i))
  for (j in 1:length(selectedGeneList)){
    PG_results_sk[[j]] <- clustSensitivity(PG_clustering_sk[[i]][[j]], true_labels_vector, as.character(1:k_cls))
    print(paste("got clustering results:", j))
  }
  PG_clustering_sens_sk[[i]]<-PG_results_sk
}

#labels_vs_clust <- data.frame(
#    PG_labels     = PG_groups_sk[[1]]
#)
#labels_vs_clust$truth_labels <- true_labels_vector[rownames(labels_vs_clust)]
#celltype_vec <- setNames(true_labels$label, true_labels$cell_id)
#labels_vs_clust$celltype <- celltype_vec[rownames(labels_vs_clust)]
#write.csv(labels_vs_clust, paste0(here(), "/vst_Data_Hold/", name, "_truth_vs_clustering.csv"))
smaller_count_matrix<-count_matrix[names(principal_genes_results$top_genes[1:100]),]
write.csv(smaller_count_matrix, paste0(here(), "/vst_Data_Hold/", name, "_top100PG_count_matrix.csv"))
```

##SK_MEANS CLUSTERING USING HVGS
```{r}
library(Seurat)
set.seed(12345)
seuratObject<- CreateSeuratObject(counts = as.matrix(count_matrix))
norm_seuratObject <- NormalizeData(seuratObject)
seuratObject<- FindVariableFeatures(norm_seuratObject, selection.method = "vst", nfeatures = length(principal_genes_results$top_genes)) #variance stabilized used as the metric here. some regression stuff is done that you need to look into.
HVGs <- VariableFeatures(seuratObject) #Vector of descending order HVGs

positions<-match(HVGs, seuratObject[["RNA"]]@meta.data$var.features)

scored_hvg<- seuratObject[["RNA"]]@meta.data$vf_vst_counts_variance.standardized[positions]
names(scored_hvg) <- HVGs
scored_hvg[is.na(scored_hvg)] <- 0
scored_hvg <- sort(decreasing = TRUE, scored_hvg)

commongenes<-intersect(names(principal_genes_results$top_genes), names(scored_hvg))

true_labels<-subset(full_true_labels, cell %in% colnames(count_matrix))
true_labels$numeric_labels <- as.numeric(factor(
  true_labels$label,
  levels = unique(true_labels$label)
))
true_labels_vector <- setNames(true_labels$numeric_labels, true_labels$cell)

# SKmeans clustering
HVG_groups_sk <- list()
HVG_clustering_sk <- list()
HVG_results_sk <- list()
HVG_clustering_sens_sk <- list()

for(i in 1:length(seeds)){
  print(paste("seed:", seeds[i]))
  for(j in seq_along(num_genes)){
    set.seed(seeds[i])
    submat<-count_matrix[HVGs[1:min(length(HVGs),as.numeric(num_genes[j]))], ]
    HVG_groups_sk[[j]]<-skmeans(x = t(log2(submat + 1)) , k = k_cls, control = list(verbose = FALSE))$cluster
    print(paste0(j, "th clustering done"))
  }
  HVG_clustering_sk[[i]]<-HVG_groups_sk
}

for (i in 1:length(seeds)){
  for (j in seq_along(num_genes)){
    HVG_results_sk[[j]] <- clustSensitivity(HVG_clustering_sk[[i]][[j]], true_labels_vector, as.character(1:k_cls))
  }
  HVG_clustering_sens_sk[[i]]<-HVG_results_sk
}

```

##SK MEANS CLUSTERING USING UNION OF HVGS AND Principal Genes AT EACH POINT
```{r}
pg_hvg_union<-list()

for(i in seq_along(selectedGeneList)){
  pg_hvg_union[[i]]<-unique(union(names(selectedGeneList[[i]]), HVGs[1:length(selectedGeneList[[i]])]))
}

# SKmeans clustering
Union_groups_sk <- list()
Union_clustering_sk<-list()
Union_results_sk <-list()
Union_clustering_sens_sk<-list()

# SKmeans clustering
for(i in 1:length(seeds)){
  print(paste("seed:", seeds[i]))
  for(j in 1:length(selectedGeneList)){
    set.seed(seeds[i])
    submat<-count_matrix[pg_hvg_union[[j]],]
    Union_groups_sk[[j]]<-skmeans(x = t(log2(submat + 1)) , k = k_cls, control = list(verbose = FALSE))$cluster
    print(paste0(j, "th clustering done"))
  }
  Union_clustering_sk[[i]]<-Union_groups_sk
}

for(i in 1:length(seeds)){
  for (j in 1:length(selectedGeneList)){
    Union_results_sk[[j]] <- clustSensitivity(Union_clustering_sk[[i]][[j]], true_labels_vector, as.character(1:k_cls))
  }
  Union_clustering_sens_sk[[i]]<-Union_results_sk
}

```

##PLOT CLUSTERING RESULTS COMPARING HVGS AND CUMULATIVE SCORE
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(mclust)

RowNames <- list()
pg_points_sk <- list()
hvg_points_sk <- list()
union_points_sk<-list()

for(i in 1: length(num_genes)){
  sens_sum=0
  for(j in 1:length(PG_clustering_sens_sk)){
    sens_sum = sens_sum + as.numeric(PG_clustering_sens_sk[[j]][[i]][1,2])
  }
  pg_points_sk[[i]] <- sens_sum/as.numeric(length(PG_clustering_sens_sk))
  RowNames[[i]] <- i
}

for(i in 1: length(num_genes)){
  sens_sum=0
  for(j in 1:length(HVG_clustering_sens_sk)){
    sens_sum = sens_sum + as.numeric(HVG_clustering_sens_sk[[j]][[i]][1,2])
  }
  hvg_points_sk[[i]] <- sens_sum/as.numeric(length(HVG_clustering_sens_sk))
}

for(i in 1: length(num_genes)){
  sens_sum=0
  for(j in 1:length(Union_clustering_sens_sk)){
    sens_sum = sens_sum + as.numeric(Union_clustering_sens_sk[[j]][[i]][1,2])
  }
  union_points_sk[[i]] <- sens_sum/as.numeric(length(Union_clustering_sens_sk))
}

points_sk <- data.frame(
  row.names = unlist(RowNames),
  PG_sensitivity=as.numeric(unlist(pg_points_sk)),
  HVG_sensitivity = as.numeric(unlist(hvg_points_sk)),
  PG_HVG_Union = as.numeric(unlist(union_points_sk)),
  thresholds =as.numeric(thresholds),
  num_genes = as.integer(num_genes),
  union_num_genes = as.integer(sapply(pg_hvg_union, length))
)

write.csv(points_sk, file=paste0(here(), "/vst_Data_Hold/TabulaMuris_Liver_SKM_accuracy_results.csv"))

# Long (tidy) format for ggplot
plot_df1 <- points_sk %>%
  pivot_longer(
    cols = c(PG_sensitivity, HVG_sensitivity, PG_HVG_Union),
    names_to = "metric",
    values_to = "sensitivity"
  )

#control legend order
plot_df1$metric <- factor(plot_df1$metric,
                         levels = c("PG_sensitivity","HVG_sensitivity","PG_HVG_Union"))

#add labels for union numbers
plot_df1$label_val <- NA
plot_df1$label_val[plot_df1$metric == "PG_sensitivity"] <- plot_df1$num_genes[plot_df1$metric == "PG_sensitivity"]
plot_df1$label_val[plot_df1$metric == "PG_HVG_Union"] <- plot_df1$union_num_genes[plot_df1$metric == "PG_HVG_Union"]

ggplot(plot_df1, aes(x = thresholds, y = sensitivity, color = metric)) +
  geom_point(size = 2) +

  # PG labels
  geom_text(
    data = subset(plot_df1, metric == "PG_sensitivity"),
    aes(label = label_val),
    color = "#1BB7D8", vjust = +2, size = 3, check_overlap = TRUE
  ) +

  # Union labels
  geom_text(
    data = subset(plot_df1, metric == "PG_HVG_Union"),
    aes(label = label_val),
    color = "darkgreen", vjust = -2, size = 3, check_overlap = TRUE
  ) +

  geom_smooth(
    method = "loess",
    method.args = list(span = 1, degree = 1, surface = "direct"),
    se = FALSE
  ) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = NULL, y = NULL, title = NULL) +
  scale_color_manual(
    breaks = c("PG_sensitivity","HVG_sensitivity","PG_HVG_Union"),
    values = c(
      "PG_sensitivity"  = "#1BB7D8",
      "HVG_sensitivity" = "#E55B1E",
      "PG_HVG_Union" = "#2C9F45"
    ),
    labels = c(
      "PG_sensitivity"  = "Principal Genes",
      "HVG_sensitivity" = "Seurat's HVG",
      "PG_HVG_Union" = "HVG_PG_Union"
    )
  ) +
  theme_minimal() + theme(legend.position = "none")

ggsave(filename = paste0(here(), "/vst_results/TabulaMuris_Liver_Results/TabulaMuris_Liver_PG_HVG_UNION_Sensitivity.jpg"),, 
     dpi = 300)

#adjustedRandIndex
pg_rand <- list()
hvg_rand <- list()
union_rand <- list()
for(i in 1:length(num_genes)){
  rand_sum_pg = 0; rand_sum_hvg = 0; rand_sum_union = 0
  for(j in 1:length(seeds)){
    rand_sum_pg    <- rand_sum_pg    + adjustedRandIndex(PG_clustering_sk[[j]][[i]],    true_labels_vector)
    rand_sum_hvg   <- rand_sum_hvg   + adjustedRandIndex(HVG_clustering_sk[[j]][[i]],   true_labels_vector)
    rand_sum_union <- rand_sum_union + adjustedRandIndex(Union_clustering_sk[[j]][[i]], true_labels_vector)
  }
  pg_rand[[i]]    <- rand_sum_pg    / length(seeds)
  hvg_rand[[i]]   <- rand_sum_hvg   / length(seeds)
  union_rand[[i]] <- rand_sum_union / length(seeds)
}

points_sk <- data.frame(
  row.names = unlist(RowNames),
  PG_randIndex=as.numeric(unlist(pg_rand)),
  HVG_randIndex = as.numeric(unlist(hvg_rand)),
  PG_HVG_UnionRand = as.numeric(unlist(union_rand)),
  thresholds =as.numeric(thresholds),
  num_genes = as.integer(num_genes),
  union_num_genes = as.integer(sapply(pg_hvg_union, length))
)
write.csv(points_sk, file=paste0(here(), "/vst_Data_Hold/TabulaMuris_Liver_SKM_rand_index_results.csv"))

# Long (tidy) format for ggplot
plot_df <- points_sk %>%
  pivot_longer(
    cols = c(PG_randIndex, HVG_randIndex, PG_HVG_UnionRand),
    names_to = "metric",
    values_to = "adjustedRandIndex"
  )

#control legend order
plot_df$metric <- factor(plot_df$metric,
                         levels = c("PG_randIndex","HVG_randIndex","PG_HVG_UnionRand"))

#add labels for union numbers
plot_df$label_val <- NA
plot_df$label_val[plot_df$metric == "PG_randIndex"] <- plot_df$num_genes[plot_df$metric == "PG_randIndex"]
plot_df$label_val[plot_df$metric == "PG_HVG_UnionRand"] <- plot_df$union_num_genes[plot_df$metric == "PG_HVG_UnionRand"]

# Plot both on the same axes

ggplot(plot_df, aes(x = thresholds, y = adjustedRandIndex, color = metric)) +
  geom_point(size = 2) +

  # PG labels
  geom_text(
    data = subset(plot_df, metric == "PG_randIndex"),
    aes(label = label_val),
    color = "#1BB7D8", vjust = +2, size = 3, check_overlap = TRUE
  ) +

  # Union labels
  geom_text(
    data = subset(plot_df, metric == "PG_HVG_UnionRand"),
    aes(label = label_val),
    color = "darkgreen", vjust = -2, size = 3, check_overlap = TRUE
  ) +

  geom_smooth(
    method = "loess",
    method.args = list(span = 1, degree = 1, surface = "direct"),
    se = FALSE
  ) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = NULL, y = NULL, title = NULL) +
  scale_color_manual(
    breaks = c("PG_randIndex","HVG_randIndex","PG_HVG_UnionRand"),
    values = c(
      "PG_randIndex"  = "#1BB7D8",
      "HVG_randIndex" = "#E55B1E",
      "PG_HVG_UnionRand" = "#2C9F45"
    ),
    labels = c(
      "PG_randIndex"  = "Principal Genes",
      "HVG_randIndex" = "Seurat's HVG",
      "PG_HVG_UnionRand" = "HVG_PG_Union"
    )
  ) +
  theme_minimal() + theme(legend.position = "none")

ggsave(filename = paste0(here(), "/vst_results/TabulaMuris_Liver_Results/TabulaMuris_Liver_PG_HVG_UNION_RandIndex.jpg"),, 
     dpi = 300)

```
##NO OUTLIERS
```{r}
add_loess_ci <- function(df, span = 1, degree = 1, surface = "direct") {
  mod <- loess(sensitivity ~ thresholds, data = df,
               span = span, degree = degree, surface = surface)
  pr  <- predict(mod, newdata = data.frame(thresholds = df$thresholds), se = TRUE)

  df$fit   <- pr$fit
  df$se    <- pr$se.fit
  df$lower <- df$fit - 1.96 * df$se
  df$upper <- df$fit + 1.96 * df$se
  df$outlier <- df$sensitivity < df$lower | df$sensitivity > df$upper
  df
}

# Apply per metric
plot_df1 <- plot_df1 %>%
  group_by(metric) %>%
  group_modify(~ add_loess_ci(.x, span = 1, degree = 1, surface = "direct")) %>%
  ungroup()

# Filter out outliers
plot_df1_filtered <- plot_df1 %>% filter(!outlier)

ggplot(plot_df1_filtered, aes(x = thresholds, y = sensitivity, color = metric)) +
  geom_point(size = 2) +

  # PG labels
  geom_text(
    data = subset(plot_df1_filtered, metric == "PG_sensitivity"),
    aes(label = label_val),
    color = "#1BB7D8", vjust = +2, size = 3, check_overlap = TRUE
  ) +

  # Union labels
  geom_text(
    data = subset(plot_df1_filtered, metric == "PG_HVG_Union"),
    aes(label = label_val),
    color = "darkgreen", vjust = -2, size = 3, check_overlap = TRUE
  ) +

  geom_smooth(
    method = "loess",
    method.args = list(span = 1, degree = 1, surface = "direct"),
    se = FALSE
  ) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = NULL, y = NULL, title = NULL) +
  scale_color_manual(
    breaks = c("PG_sensitivity","HVG_sensitivity","PG_HVG_Union"),
    values = c(
      "PG_sensitivity"  = "#1BB7D8",
      "HVG_sensitivity" = "#E55B1E",
      "PG_HVG_Union"    = "#2C9F45"
    ),
    labels = c(
      "PG_sensitivity"  = "Principal Genes",
      "HVG_sensitivity" = "Seurat's HVG",
      "PG_HVG_Union"    = "HVG_PG_Union"
    )
  ) +
  theme_minimal() + theme(legend.position = "none")

ggsave(filename = paste0(here(), "/vst_results/TabulaMuris_Liver_Results/TabulaMuris_Liver_PG_HVG_UNION_No_Outliers_Sensitivity.jpg"),, 
     dpi = 300)

add_loess_ci <- function(df, span = 1, degree = 1, surface = "direct") {
  mod <- loess(adjustedRandIndex ~ thresholds, data = df,
               span = span, degree = degree, surface = surface)
  pr  <- predict(mod, newdata = data.frame(thresholds = df$thresholds), se = TRUE)

  df$fit   <- pr$fit
  df$se    <- pr$se.fit
  df$lower <- df$fit - 1.96 * df$se
  df$upper <- df$fit + 1.96 * df$se
  df$outlier <- df$adjustedRandIndex < df$lower | df$adjustedRandIndex > df$upper
  df
}

# Apply per metric
plot_df <- plot_df %>%
  group_by(metric) %>%
  group_modify(~ add_loess_ci(.x, span = 1, degree = 1, surface = "direct")) %>%
  ungroup()

# Filter out outliers
plot_df_filtered <- plot_df %>% filter(!outlier)

ggplot(plot_df_filtered, aes(x = thresholds, y = adjustedRandIndex, color = metric)) +
  geom_point(size = 2) +

  # PG labels
  geom_text(
    data = subset(plot_df_filtered, metric == "PG_randIndex"),
    aes(label = label_val),
    color = "#1BB7D8", vjust = +2, size = 3, check_overlap = TRUE
  ) +

  # Union labels
  geom_text(
    data = subset(plot_df_filtered, metric == "PG_HVG_UnionRand"),
    aes(label = label_val),
    color = "darkgreen", vjust = -2, size = 3, check_overlap = TRUE
  ) +

  geom_smooth(
    method = "loess",
    method.args = list(span = 1, degree = 1, surface = "direct"),
    se = FALSE
  ) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = NULL, y = NULL, title = NULL) +
  scale_color_manual(
    breaks = c("PG_randIndex","HVG_randIndex","PG_HVG_UnionRand"),
    values = c(
      "PG_randIndex"  = "#1BB7D8",
      "HVG_randIndex" = "#E55B1E",
      "PG_HVG_UnionRand"    = "#2C9F45"
    ),
    labels = c(
      "PG_randIndex"  = "Principal Genes",
      "HVG_randIndex" = "Seurat's HVG",
      "PG_HVG_UnionRand"    = "HVG_PG_Union"
    )
  ) +
  theme_minimal() + theme(legend.position = "none")

ggsave(filename = paste0(here(), "/vst_results/TabulaMuris_Liver_Results/TabulaMuris_Liver_PG_HVG_UNION_No_Outliers_Sensitivity.jpg"),, 
     dpi = 300)

```

##PRINCIPAL GENES AND HVGS COMPARISON
```{r}
#Common genes between TOP 100 Principal Genes results and Highly Variable Genes
common_gene_names<- intersect(names(principal_genes_results$top_genes[1:100]), names(scored_hvg[1:100]))
pg_only_gene_names <- setdiff(names(principal_genes_results$top_genes[1:100]), common_gene_names)
hvg_only_gene_names <-setdiff(names(scored_hvg[1:100]), common_gene_names)

#Ranks of Principal Genes based on scores
pps_ranks <- rank(-as.numeric(principal_genes_results$top_genes))
names(pps_ranks) <- names(principal_genes_results$top_genes)

#Ranks of HVGs based on scaled dispersion score
hvg_ranks <- rank(-as.numeric(scored_hvg))
names(hvg_ranks) <- names(scored_hvg)

#Variance of each gene
variance_values <- apply(count_matrix, 1, var, na.rm = TRUE)
variance_values <- sort(decreasing = TRUE, variance_values)

#Mean expression of each gene
mean_values <- apply(count_matrix, 1, mean, na.rm = TRUE)
mean_values <- sort(decreasing = TRUE, mean_values)

#Compare the ranks of intersecting genes
intersect_df <- data.frame(
  PPS_rank = pps_ranks[common_gene_names],
  HVG_rank = hvg_ranks[common_gene_names],
  Variance = round(variance_values[common_gene_names], 4),
  Mean = round(mean_values[common_gene_names], 4)
)


#Compare the scores of the top100 Principal Genes to their rank in the top 100 HVGs and their mean and variance scores
raw_scores_df <- data.frame(
  PPS_score_percent = principal_genes_results$top_genes[1:100]*100,
  HVG_rank = hvg_ranks[names(principal_genes_results$top_genes[1:100])],
  Variance = round(variance_values[names(principal_genes_results$top_genes[1:100])], 4),
  Mean = round(mean_values[names(principal_genes_results$top_genes[1:100])], 4)
)

write.csv(intersect_df, file = paste0(here(), "/vst_Data_Hold/TabulaMuris_Liver_PPS_HVG_Variance_Mean_Intersect.csv"))
write.csv(raw_scores_df, file = paste0(here(), "/vst_Data_Hold/TabulaMuris_Liver_PPS_HVG_Variance_Mean_Top100.csv"))
```

##FINAL PLOTS FROM HVG-only genes and PPS-only genes
```{r}
library(dplyr)
library(ggplot2)
library(ggrepel)

#plot Stabilized Variance vs GPS
genes <- rownames(count_matrix)

graphdf <- data.frame(
  genes = genes,
  stabilized_variance = scored_hvg[match(genes, names(scored_hvg))],
  GPS = principal_genes_results$top_genes[match(genes, names(principal_genes_results$top_genes))]
)


#Compare GPS and Stabilized Variance
pg20   <- head(pg_only_gene_names[pg_only_gene_names %in% graphdf$genes], 20)
hvg20  <- head(hvg_only_gene_names[hvg_only_gene_names %in% graphdf$genes], 20)
both20 <- head(common_gene_names[common_gene_names %in% graphdf$genes], 20)

# Build comparison df with categories
comparison_df <- graphdf %>%
  mutate(
    category = case_when(
      genes %in% pg20   ~ "PG only",
      genes %in% hvg20  ~ "HVG only",
      genes %in% both20 ~ "Both PG & HVG",
      TRUE              ~ "Other"
    ),
    category = factor(category, levels = c("PG only", "HVG only", "Both PG & HVG", "Other"))
  ) %>%
  filter(is.finite(GPS), is.finite(stabilized_variance))

# Separate for background vs highlights
other_df     <- comparison_df %>% filter(category == "Other")
highlight_df <- comparison_df %>% filter(category != "Other")

# Plot: GPS vs Stabilized variance with highlighted top-20s
ggplot() +
  # Background points
  geom_point(
    data = other_df,
    aes(x = GPS, y = stabilized_variance),
    color = "grey70", alpha = 0.4, size = 1.6
  ) +
  # Highlighted categories
  geom_point(
    data = highlight_df,
    aes(x = GPS, y = stabilized_variance, color = category),
    alpha = 0.7, size = 1.8
  ) +
  # Labels for highlighted genes
  ggrepel::geom_text_repel(
    data = highlight_df,
    aes(x = GPS, y = stabilized_variance, label = genes, color = category),
    size = 2.5, max.overlaps = Inf, show.legend = FALSE
  ) +
  labs(x = NULL, y = NULL, title = NULL) +
  theme_minimal() + theme(legend.position = "none") +
  scale_color_manual(
    values = c(
      "PG only"       = "#1BB7D8",
      "HVG only"      = "#E55B1E",
      "Both PG & HVG" = "#2C9F45"
    )
  ) +
  theme(plot.title = element_text(hjust = 0.5))
ggsave(filename = paste0(here(), "/vst_results/TabulaMuris_Liver_Results/TabulaMuris_Liver_GPS_Stabilized_Variance_Comparison.jpg"),, 
     dpi = 300)

# Build comparison_df with all categories
comparison_df <- log_var_mean_df %>%
  mutate(
    category = case_when(
      gene_id %in% pg_only_gene_names[1:20]   ~ "PG only",
      gene_id %in% hvg_only_gene_names[1:20]  ~ "HVG only",
      gene_id %in% common_gene_names[1:20]    ~ "Both PG & HVG",
      TRUE ~ "Other"
    )
  )

# Separate dataframes
other_df   <- comparison_df %>% filter(category == "Other")
highlight_df <- comparison_df %>% filter(category != "Other")

# Plot
ggplot() +
  # Background grey points (no legend)
  geom_point(
    data = other_df,
    aes(x = log_mean, y = log_var),
    color = "grey70", alpha = 0.4, size = 1.6, inherit.aes = FALSE
  ) +
  # Highlighted points with legend
  geom_point(
    data = highlight_df,
    aes(x = log_mean, y = log_var, color = category),
    alpha = 0.7, size = 1.8
  ) +
  # Labels for all highlight genes
  ggrepel::geom_text_repel(
    data = highlight_df,
    aes(x = log_mean, y = log_var, label = gene_id, color = category),
    size = 2.5, max.overlaps = Inf, show.legend = FALSE
  ) +
  labs(x = NULL, y = NULL, title = NULL) +
  theme_minimal() + theme(legend.position = "none") +
  scale_color_manual(
    values = c(
      "PG only"       = "#1BB7D8",
      "HVG only"      = "#E55B1E",
      "Both PG & HVG" = "#2C9F45"
    )
  ) +
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "red")

ggsave(filename = paste0(here(), "/vst_results/TabulaMuris_Liver_Results/TabulaMuris_Liver_PG_HVG_INTERSECT_Log10_mean_variance.jpg"),, 
     dpi = 300)

##VIOLIN PLOT
## VIOLIN PLOT

# PG Only
tryCatch({
  VlnPlot(norm_seuratObject, features = pg_only_gene_names[1:9]) & 
    theme(axis.title.x = element_blank(),
          axis.text.x = element_blank(),
          axis.title.y = element_blank(),
          legend.position = "none")
  ggsave(filename = paste0(here(), "/vst_results/TabulaMuris_Liver_Results/TabulaMuris_Liver_PG_Only_Violin_Plot.jpg"),
         dpi = 300)
}, error = function(e) {
  message("PG Only Violin Plot failed: ", e$message)
})

# HVG Only
tryCatch({
  VlnPlot(norm_seuratObject, features = hvg_only_gene_names[1:9]) & 
    theme(axis.title.x = element_blank(),
          axis.text.x = element_blank(),
          axis.title.y = element_blank(),
          legend.position = "none")
  ggsave(filename = paste0(here(), "/vst_results/TabulaMuris_Liver_Results/TabulaMuris_Liver_HVG_Only_Violin_Plot.jpg"),
         dpi = 300)
}, error = function(e) {
  message("HVG Only Violin Plot failed: ", e$message)
})

# PG + HVG Intersect
tryCatch({
  VlnPlot(norm_seuratObject, features = common_gene_names[1:9]) & 
    theme(axis.title.x = element_blank(),
          axis.text.x = element_blank(),
          axis.title.y = element_blank(),
          legend.position = "none")
  ggsave(filename = paste0(here(), "/vst_results/TabulaMuris_Liver_Results/TabulaMuris_Liver_PG_HVG_INTERSECT_Violin_Plot.jpg"),
         dpi = 300)
}, error = function(e) {
  message("PG+HVG Intersect Violin Plot failed: ", e$message)
})
```
